---
title: "Supplementary tables for the article \"Model-comparison effect size indices
  for logistic and other generalized linear models: Computation and power analysis\""
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulation results tables

The figures in the main text plot the results of Study 1 (recovery of the whole-model
R^2^ and of a single focal predictor's eta-squared) and Study 2 (simulation-based
required sample size). For Study 1, each Model x N x k combination in the figures is a
curve over 10 population effect size values; the tables below summarize accuracy and
power collapsed across those population effect size values only (mean absolute bias,
RMSE, and mean power gap, by model, sample size, and number of covariates k), keeping
the k = 3 vs. k = 5 conditions separate rather than averaging over them, so the full
crossed design is preserved -- the full per-point results are available in this
repository's `simulations/Data/` folder. For Study 2, which has no sample-size or k
dimension, the tables report the full underlying data.

```{r results-setup, include=FALSE}
library(kableExtra)
source("../simulations/functions.R")   # theoretical_power(), focal_df(), model_labels

load("../simulations/Data/study1.a.data.Rdata")   # -> `results` (Study 1a, R^2 design)
study1a <- results

load("../simulations/Data/study1.b.data.Rdata")   # -> `results` (Study 1b, eta^2 design)
study1b <- results

e <- new.env(); load("../simulations/Data/study2.a.data.Rdata", envir = e); study2a <- e$results
e <- new.env(); load("../simulations/Data/study2.b.data.Rdata", envir = e); study2b <- e$results

study1a$power_r2    <- mapply(theoretical_power, study1a$er2,  study1a$N, study1a$k, study1a$model)
study1a$power_adjr2 <- mapply(theoretical_power, study1a$aer2, study1a$N, study1a$k, study1a$model)

study1b$df_test <- sapply(study1b$model, focal_df)
suppressMessages({
  study1b$power_eta2    <- mapply(theoretical_power, study1b$eeta2,  study1b$N, study1b$k, study1b$model, study1b$df_test)
  study1b$power_adjeta2 <- mapply(theoretical_power, study1b$aeeta2, study1b$N, study1b$k, study1b$model, study1b$df_test)
})

long_table <- function(data, cols, names, order_cols, digits, caption) {
  data$Model <- factor(model_labels[data$model], levels = model_labels)
  tab <- data[do.call(order, data[, c("Model", order_cols)]), c("Model", cols)]
  colnames(tab) <- c("Model", names)
  rownames(tab) <- NULL
  kbl(tab, digits = digits, booktabs = TRUE, longtable = TRUE, caption = caption,
      linesep = c(rep("", 7), "\\addlinespace")) %>%
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 8)
}

## Collapse a Study 1 data set across its 10 population-effect-size values only, giving
## one row per Model x N x k cell -- 32 rows total, vs. 320 in the full per-point data.
## k is kept as its own column (not averaged over), so the full crossed design shows.
agg_table <- function(data, formula, names, digits, caption) {
  data$model_label <- factor(model_labels[data$model], levels = model_labels)
  tab <- aggregate(formula, data = data, FUN = mean)
  tab <- tab[order(tab$model_label, tab$N, tab$k), ]
  colnames(tab) <- c("Model", "N", "k", names)
  rownames(tab) <- NULL
  kbl(tab, digits = digits, booktabs = TRUE, longtable = TRUE, caption = caption,
      linesep = c(rep("", 7), "\\addlinespace")) %>%
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 9)
}

study1a$power_gap_r2    <- abs(study1a$power_r2    - study1a$power)
study1a$power_gap_adjr2 <- abs(study1a$power_adjr2 - study1a$power)

study1b$power_gap_eta2 <- abs(study1b$power_eta2    - study1b$power)
study1b$power_gap_eps2 <- abs(study1b$power_adjeta2 - study1b$power)
```

### Study 1a: accuracy of R^2^ (collapsed summary of Figure 1)

```{r table-figure1, echo=FALSE}
agg_table(study1a,
          cbind(abs_bias_r2 = abs(bias), rmse_r2 = rmse,
                abs_bias_adjr2 = abs(abias), rmse_adjr2 = armse) ~ model_label + N + k,
          c("Mean |Bias| (R2)", "RMSE (R2)", "Mean |Bias| (Adjusted R2)", "RMSE (Adjusted R2)"),
          digits = 3,
          caption = "Accuracy of R2 and adjusted R2, by model, N, and number of covariates k, collapsed across population R2 (data underlying Figure 1).")
```

### Study 1a: power of the omnibus test (collapsed summary of Figure 3)

```{r table-figure3, echo=FALSE}
agg_table(study1a,
          cbind(power_gap_r2, power_gap_adjr2) ~ model_label + N + k,
          c("Mean Power Gap (R2)", "Mean Power Gap (Adjusted R2)"),
          digits = 3,
          caption = "Mean |simulated - closed-form| power for R2 and adjusted R2, by model, N, and number of covariates k, collapsed across population R2 (data underlying Figure 3).")
```

### Study 1b: accuracy of eta-squared (collapsed summary of Figure 2)

```{r table-figure2, echo=FALSE}
agg_table(study1b,
          cbind(abs_bias_eta2 = abs(bias), rmse_eta2 = rmse,
                abs_bias_eps2 = abs(abias), rmse_eps2 = armse) ~ model_label + N + k,
          c("Mean |Bias| (Eta2)", "RMSE (Eta2)", "Mean |Bias| (Epsilon2)", "RMSE (Epsilon2)"),
          digits = 3,
          caption = "Accuracy of eta-squared and adjusted epsilon-squared, by model, N, and number of covariates k, collapsed across population eta-squared (data underlying Figure 2).")
```

### Study 1b: power of the single-predictor test (collapsed summary of Figure 4)

```{r table-figure4, echo=FALSE}
agg_table(study1b,
          cbind(power_gap_eta2, power_gap_eps2) ~ model_label + N + k,
          c("Mean Power Gap (Eta2)", "Mean Power Gap (Epsilon2)"),
          digits = 3,
          caption = "Mean |simulated - closed-form| power for eta-squared and epsilon-squared, by model, N, and number of covariates k, collapsed across population eta-squared (data underlying Figure 4).")
```

### Study 2: required sample size (data underlying Figure 5)

```{r table-figure5-omnibus, echo=FALSE}
long_table(study2a, c("r2", "N", "Nt"),
           c("Population R2", "Required N (Simulated)", "Required N (Closed-form)"),
           order_cols = "r2", digits = 3,
           caption = "Required sample size for 80\\% power, omnibus test: simulated vs. closed-form.")
```

```{r table-figure5-individual, echo=FALSE}
long_table(study2b, c("eta2", "N", "Nt"),
           c("Population Eta2", "Required N (Simulated)", "Required N (Closed-form)"),
           order_cols = "eta2", digits = 3,
           caption = "Required sample size for 80\\% power, single-predictor test: simulated vs. closed-form.")
```
