---
title: "Supplementary material for the article \"Model-comparison effect size indices
  for logistic and other generalized linear models: Computation and power analysis\""
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulation results tables

The figures in the main text plot the results of Study 1 (recovery of the whole-model
R^2^ and of a single focal predictor's eta-squared) and Study 2 (simulation-based
required sample size). For Study 1, each Model x N combination in the figures is a curve
over 16 population effect size values; the tables below summarize accuracy and power
collapsed across those 16 values (mean absolute bias, RMSE, and mean power gap, by model
and sample size), rather than reproducing every plotted point -- the full per-point
results are available in this repository's `simulations/Data/` folder. For Study 2, which
has no sample-size dimension, the tables report the full underlying data.

```{r results-setup, include=FALSE}
library(kableExtra)
source("../simulations/functions.R")   # theoretical_power(), focal_df(), model_labels

load("../simulations/Data/study1.a.data.Rdata")   # -> `results` (Study 1a, R^2 design)
study1a <- results

load("../simulations/Data/study1.b.data.Rdata")   # -> `results` (Study 1b, eta^2 design)
study1b <- results

e <- new.env(); load("../simulations/Data/study2.a.data.Rdata", envir = e); study2a <- e$results
e <- new.env(); load("../simulations/Data/study2.b.data.Rdata", envir = e); study2b <- e$results

study1a$power_r2    <- mapply(theoretical_power, study1a$er2,  study1a$N, study1a$k, study1a$model)
study1a$power_adjr2 <- mapply(theoretical_power, study1a$aer2, study1a$N, study1a$k, study1a$model)

study1b$df_test <- sapply(study1b$model, focal_df)
suppressMessages({
  study1b$power_eta2    <- mapply(theoretical_power, study1b$eeta2,  study1b$N, study1b$k, study1b$model, study1b$df_test)
  study1b$power_adjeta2 <- mapply(theoretical_power, study1b$aeeta2, study1b$N, study1b$k, study1b$model, study1b$df_test)
})

long_table <- function(data, cols, names, order_cols, digits, caption) {
  data$Model <- model_labels[data$model]
  tab <- data[do.call(order, data[, c("Model", order_cols)]), c("Model", cols)]
  colnames(tab) <- c("Model", names)
  rownames(tab) <- NULL
  kbl(tab, digits = digits, booktabs = TRUE, longtable = TRUE, caption = caption,
      linesep = c(rep("", 7), "\\addlinespace")) %>%
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 8)
}

## Collapse a Study 1 data set across its 16 population-effect-size values, giving one
## row per Model x N cell -- 16 rows total, vs. 256 in the full per-point data.
agg_table <- function(data, formula, names, digits, caption) {
  data$model_label <- model_labels[data$model]
  tab <- aggregate(formula, data = data, FUN = mean)
  tab <- tab[order(tab$model_label, tab$N), ]
  colnames(tab) <- c("Model", "N", names)
  rownames(tab) <- NULL
  kbl(tab, digits = digits, booktabs = TRUE, longtable = TRUE, caption = caption,
      linesep = c(rep("", 3), "\\addlinespace")) %>%
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 9)
}

study1a$power_gap_r2    <- abs(study1a$power_r2    - study1a$power)
study1a$power_gap_adjr2 <- abs(study1a$power_adjr2 - study1a$power)

study1b$power_gap_eta2 <- abs(study1b$power_eta2    - study1b$power)
study1b$power_gap_eps2 <- abs(study1b$power_adjeta2 - study1b$power)
```

### Study 1a: accuracy of R^2^ (collapsed summary of Figure 1)

```{r table-figure1, echo=FALSE}
agg_table(study1a,
          cbind(abs_bias_r2 = abs(bias), rmse_r2 = rmse,
                abs_bias_adjr2 = abs(abias), rmse_adjr2 = armse) ~ model_label + N,
          c("Mean |Bias| (R2)", "RMSE (R2)", "Mean |Bias| (Adjusted R2)", "RMSE (Adjusted R2)"),
          digits = 3,
          caption = "Accuracy of R2 and adjusted R2, collapsed across population R2 (data underlying Figure 1).")
```

### Study 1a: power of the omnibus test (collapsed summary of Figure 3)

```{r table-figure3, echo=FALSE}
agg_table(study1a,
          cbind(power_gap_r2, power_gap_adjr2) ~ model_label + N,
          c("Mean Power Gap (R2)", "Mean Power Gap (Adjusted R2)"),
          digits = 3,
          caption = "Mean |simulated - closed-form| power for R2 and adjusted R2, collapsed across population R2 (data underlying Figure 3).")
```

### Study 1b: accuracy of eta-squared (collapsed summary of Figure 2)

```{r table-figure2, echo=FALSE}
agg_table(study1b,
          cbind(abs_bias_eta2 = abs(bias), rmse_eta2 = rmse,
                abs_bias_eps2 = abs(abias), rmse_eps2 = armse) ~ model_label + N,
          c("Mean |Bias| (Eta2)", "RMSE (Eta2)", "Mean |Bias| (Epsilon2)", "RMSE (Epsilon2)"),
          digits = 3,
          caption = "Accuracy of eta-squared and adjusted epsilon-squared, collapsed across population eta-squared (data underlying Figure 2).")
```

### Study 1b: power of the single-predictor test (collapsed summary of Figure 4)

```{r table-figure4, echo=FALSE}
agg_table(study1b,
          cbind(power_gap_eta2, power_gap_eps2) ~ model_label + N,
          c("Mean Power Gap (Eta2)", "Mean Power Gap (Epsilon2)"),
          digits = 3,
          caption = "Mean |simulated - closed-form| power for eta-squared and epsilon-squared, collapsed across population eta-squared (data underlying Figure 4).")
```

### Study 2: required sample size (data underlying Figure 5)

```{r table-figure5-omnibus, echo=FALSE}
long_table(study2a, c("r2", "N", "Nt"),
           c("Population R2", "Required N (Simulated)", "Required N (Closed-form)"),
           order_cols = "r2", digits = 3,
           caption = "Required sample size for 80\\% power, omnibus test: simulated vs. closed-form.")
```

```{r table-figure5-individual, echo=FALSE}
long_table(study2b, c("eta2", "N", "Nt"),
           c("Population Eta2", "Required N (Simulated)", "Required N (Closed-form)"),
           order_cols = "eta2", digits = 3,
           caption = "Required sample size for 80\\% power, single-predictor test: simulated vs. closed-form.")
```

## Supplementary code: generating a logistic-regression sample with a known population R-squared

### Purpose

This file illustrates, for a single case -- a logistic regression whose population
R-squared is exactly known in advance -- the calibrate-then-sample method used throughout
the simulations reported in the paper. The code below is a stripped-down extract of the
logistic branch of `sample_by_r2()` from the `Rsimcity` package
(<https://github.com/mcfanda/RsimCity>): input validation, caching, and the branches for
the other model families (gaussian, multinomial, ordinal) and for targeting a single
predictor's eta-squared rather than the whole model's R-squared have all been removed, so
that only the logic essential to this one case remains. All of those variations, their
defaults, and their full implementations can be found in the package itself.

### The method in three steps

1. Build a numerical-integration grid over the joint distribution of the `k` predictors,
   using Gauss-Hermite quadrature.
2. Find, by root-finding, the coefficient scale that makes the *population* R-squared --
   computed exactly on that grid -- equal the requested target.
3. Only then draw an actual finite sample of size `n` from the calibrated population.

### Code

```{r sample-logistic-r2}
sample_logistic_r2 <- function(n, R2, k = 1, beta = rep(1, k),
                                intercept = 0, x_mean = rep(0, k), x_cov = diag(k),
                                order = 10, seed = NULL) {

  ## Gauss-Hermite quadrature grid over the k predictors
  ## using statmod::gauss.quad.prob(). It returns nodes/weights for a standard normal density,
  ## with weights already summing to 1, so that for any function f, E[f(X)] = sum(w * f(Xgh)).
  ## The grid is then transformed through the Cholesky factor of x_cov and shifted by x_mean, giving quadrature nodes
  ## for the actual (correlated, possibly non-centered) predictor distribution.
  gh <- statmod::gauss.quad.prob(order, dist = "normal")
  g  <- as.matrix(expand.grid(rep(list(seq_len(order)), k)))
  Z  <- matrix(gh$nodes[g], ncol = k)
  w  <- apply(matrix(gh$weights[g], ncol = k), 1, prod)
  Xgh <- sweep(Z %*% chol(x_cov), 2, x_mean, "+")

  ## Population deviance R-squared, as a function of a coefficient scale `a`
  ## beta = a * beta direction; the model's probability at every grid node gives the
  ## deviance-based R-squared exactly (no simulation is involved in this computation).
  r2_of_a <- function(a) {
    eta  <- intercept + drop(Xgh %*% (a * beta))
    prob <- stats::plogis(eta)

    p    <- cbind(1 - prob, prob)
    p    <- pmax(p, .Machine$double.eps)
    p0   <- colSums(p * w)                        # population marginal probabilities
    D1   <- -2 * sum(w * rowSums(p * log(p)))      # model deviance
    D0   <- -2 * sum(p0 * log(p0))                 # null-model deviance
    1 - D1 / D0
  }

  ## ---- 3. Root-find the coefficient scale that produces the target R2 ----------------
  upper <- 1
  while (r2_of_a(upper) < R2 && upper < 1e6) upper <- upper * 2
  a <- stats::uniroot(function(a) r2_of_a(a) - R2, c(0, upper))$root
  beta_final <- a * beta

  ## ---- 4. Draw the actual finite sample from the calibrated population ---------------
  if (!is.null(seed)) set.seed(seed)
  X <- sweep(matrix(stats::rnorm(n * k), n, k) %*% chol(x_cov), 2, x_mean, "+")
  colnames(X) <- paste0("x", seq_len(k))
  eta <- intercept + drop(X %*% beta_final)
  y   <- stats::rbinom(n, 1, stats::plogis(eta))

  data.frame(y = y, X)
}
```

### Example: a single predictor, R-squared = .10

```{r example}
dat <- sample_logistic_r2(n = 2000, R2 = .10, seed = 1)
mod <- glm(y ~ x1, data = dat, family = binomial())
1 - mod$deviance / mod$null.deviance   # sample deviance R2, close to the .10 target
```

### Example: three correlated predictors, R-squared = .20

```{r example-multi}
Sigma <- matrix(c(1, .3, .2,
                   .3, 1, .4,
                   .2, .4, 1), nrow = 3)

dat3 <- sample_logistic_r2(n = 5000, R2 = .20, k = 3, x_cov = Sigma, seed = 1)
mod3 <- glm(y ~ x1 + x2 + x3, data = dat3, family = binomial())
1 - mod3$deviance / mod3$null.deviance   # close to the .20 target
```
